Using SMT engine to generate Symbolic Automata\thanks{This work was partially funded by the Associated Team FM4CPS between INRIA and ECNU, Shanghai}

### 2018-01-22 2017-Xudong-SMT/FM-2018-extended.tex
Some edits and shortening

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet pn (cannot be reduce to a hole)
\State Initialize sets $U=\{\mathit{InitState}(pn)\}$, $E=\emptyset$
for unexplored/explored global states, $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in $U$; add $S$ in $E$;
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\ForAll{$\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& $T \not\in E$)}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(pn),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule (see TR1 in
Appendix~\ref{appendix:semRules}) to build the OT. 
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as detailed in algorithm~\ref{alg2}. This method also manages directly the
holes of the node, so \emph{MakeTransitions} will never be called on a Hole.

Then, at the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability. Unsatisfiable
OTs are pruned, and the final open automaton is computed as the set of
global states reachable by satisfiable OTs.

@@

\noteSB{Line 7: There are two type mismatches here: $ot$ is an open
  transition, whereas $\mathit{Matching}$ is a predicate\mdash one
  cannot assign a predicate to an open transition.  Furthermore, in
  the definition of $\mathit{Matching}$, below, $C_n$ is an open
  transition, whereas $\overline{comb}$ is a {\em set} of open
  transitions, so this call does not type either. There is probably
  another inner loop over all tuples of OTs in $\overline{comb}$; even
  so the open transition has to be defined completely, not just the
  predicate\mdash what is the value of $Post$ for the new transition?
  Maybe $\mathit{Matching}$ should not be a predicate, but an OT?
  Finally, how is $\mathit{hole}$? (I have removed $v$\mdash here and
  in the parameters of the definition of the $\mathit{Matching}$
  predicate, because the text in the paragraph ``Matching'' says that
  $v$ is fresh.)}
\noteEM{Done, please check the new version}

@@

\begin{algorithm}
\caption{MakeTransitions() \em{for a pNet node}}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $Pn$ with subnets $\overline{sn}$ and holes $\overline{hole}$; a global state $S$.
\State Initialize empty list $l$ and set $L$ for sub-transitions/transitions.
	\For{each $\mathit{Subnet}$ in $\overline{sn}$}
	~~~ $\setminus \setminus$ Recursively applying the semantic rules on the sub-nodes.
                \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\State $\overline{comb} = \mathit{Combining}(l)$;
	\For{ each $sv \in SV$, for each $comb \in \overline{comb}$ }
           \State $ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \overline{\mathit{hole}})$;
           \State Store $ot$ in $L$; ~~~ $\setminus \setminus$ when $\mathit{Matching}(ot)$ succeeds
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets, as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet; %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of n-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$, where, for each $i \in [1, n]$, $\overline{ot}_i$ is the set of open transitions obtained from the $i$-th subnet in line~3 of the algorithm.

@@

\paragraph{Matching.}
The \emph{Matching} method builds the OTs of a pNet node from those of
its subnets (see rule Tr2 in Appendix~\ref{appendix:semRules}).
For each synchronisation vector, and each possible
combination of behaviour of the subnets as generated by the \emph
{Combining} method, it builds the open transition.
By lack of space, we only detail here the construction of the predicate.
From a synchronization vector
$sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$
and its guard $G_k$; a tuple of open transitions
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$, such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$; the hole behaviors $\mathit{Hole} = (b_j)^{j \in J}$; and a fresh variable $v$, representing the result action of the OT under construction, we build the predicate:
%
\begin{multline*}
  \mathit{MkPred}(sv, C, \overline{\mathit{hole}}) = 
  \\
  (\forall i\in I, a_i=a'_i) \land
  (\forall i \not\in I, ot_i = -) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k.
\end{multline*}

@@

\begin{figure}[h]
\begin{mathpar}
    ot = \openrule{\{s0 \xrightarrow{fail(true)} s0, t0 \xrightarrow{resume(false)} t0\}, ~~\{\xrightarrow{hb}\},~~fail(true)=fail(b_1)\\
    \OTland resume(false)=start(b_2)\OTland hb=fail(b_0)\OTland v=\underline{fail}\OTland b_1=b_2\\
    \OTland (b_1\lor b_2) \Rightarrow b_0, ~~\{\}}
    	   {\ostate{s0,t0} \xrightarrow{v} \ostate{s0,t0}}
  \end{mathpar}
\caption{One of the unsatisfiable open transitions in the Failure Monitor pNet}  \label{schema:unsat-ot}
\end{figure}

@@

\begin{figure}[h]
  \centerline{\includegraphics[width=12cm]{XFIG/FailureTimerOA}}
  \caption{Open Automaton for the Failure Monitor architecture}
  \label{schema:resultOA}
\end{figure}

@@

In this paper, we presented an algorithm for generating an open
automaton representing the semantics of an open pNet.
We have also presented our implementation, which comprises two main parts. Firstly, we
compute all the possible open transitions.
\hl{The algorithm combines the actions of subnets and
hole behaviors, which are then matched with the
synchronization vectors by the generated
predicate.}  Some of the open transitions obtained at this
intermediate step, constructed in a structural manner from all
possible combinations of possible logical predicates, may
contain predicates which do not represent any possible concrete
instantiations. 
Secondly, to get rid of these useless transitions, we use the SMT solver Z3 for checking the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. We implemented our algorithm in the VerCors platform and 
tested the algorithm on available examples, such as encoding expressions from various process algebras. Our case-studies show that our encoding
successfully identifies all unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.

>>>>

\begin{algorithm}[h]
  \caption{Open Automaton Generation}
  \label{alg1}
\begin{algorithmic}[1]
\Require The pNet \hl{$P$} (cannot be a hole)
        \qquad \hl{\textbackslash\textbackslash~SB: was `` pn ''}
\State Initialize sets $U=\{\mathit{InitState}(pn)\}$ and $E=\emptyset$,
for unexplored and explored global states, respectively; $L=\emptyset$ for the resulting OTs;
\While{!$\mathit{isEmpty}(U)$}
	\State Choose $S$ in \hl{$U\setminus E$}; add $S$ in $E$;
        \qquad \hl{\textbackslash\textbackslash~SB: was `` Choose $S$ in $U$ ''}
	\State $\mathit{OT} = \mathit{MakeTransitions}(P, S)$;
        \State Store $\mathit{OT}$ in $L$; 
	\For {each $\mathit{OT} \in L$}
        \State Let $T$ be the target global state of $\mathit{OT}$
	\IfThen{($T \not\in U$ \&\& $T \not\in E$)}{Add $T$ into $U$;}
	\EndFor
\EndWhile
\State Prune the unsatisfiable open transitions in $L$ using the SMT solver;
\State \Return $\mathit{OA}=\mathit{Reachable}(\mathit{InitState}(pn),L)$;

@@

The inside loop (\emph{MakeTransitions} method) applies recursively
the semantic rules following the structure of the pNet.
When applied to a pLTSs at the leaves, we simply take the pLTS transitions of the
corresponding local state and use the semantic rule (see TR1 in
App.~\ref{appendix:semRules}) to build the OT.\footnote{%
%
  \hl{We omit detailed presentation of this case for the sake of brevity.}
%
}
When applied to a pNet node we use two methods, combining and
matching, to generate the open transitions in a hierarchical manner,
as shown in Alg.~\ref{alg2}. This method directly manages the
holes of the node, so \emph{MakeTransitions} is never called on a hole.

At the root of the pNet, the predicate of each OT is translated
into SMTlib assertions, and checked for satisfiability.
%
The final open automaton is computed as the set of global states,
reachable after pruning the unsatisfiable OTs.
%
%% Unsatisfiable
%% OTs are pruned, and the final open automaton is computed as the set of
%% global states reachable by satisfiable OTs.

@@

%% \noteSB{Line 7: There are two type mismatches here: $ot$ is an open
%%   transition, whereas $\mathit{Matching}$ is a predicate\mdash one
%%   cannot assign a predicate to an open transition.  Furthermore, in
%%   the definition of $\mathit{Matching}$, below, $C_n$ is an open
%%   transition, whereas $\overline{comb}$ is a {\em set} of open
%%   transitions, so this call does not type either. There is probably
%%   another inner loop over all tuples of OTs in $\overline{comb}$; even
%%   so the open transition has to be defined completely, not just the
%%   predicate\mdash what is the value of $Post$ for the new transition?
%%   Maybe $\mathit{Matching}$ should not be a predicate, but an OT?
%%   Finally, how is $\mathit{hole}$? (I have removed $v$\mdash here and
%%   in the parameters of the definition of the $\mathit{Matching}$
%%   predicate, because the text in the paragraph ``Matching'' says that
%%   $v$ is fresh.)}
%% \noteEM{Done, please check the new version}

@@

\begin{algorithm}
\caption{MakeTransitions() \em{for a pNet node}}
  \label{alg2}
\begin{algorithmic}[1]
\Require a pNet node $P$ with subnets $\overline{sn}$ and holes $\overline{hole}$; a global state $S$.
\State Initialize empty list $l$ and set $L$ for sub-transitions and transitions, respectively;
	\For{each $\mathit{Subnet}$ in $\overline{sn}$}
        \hfill $\setminus \setminus$ Recursively apply the semantic rules on the subnets
           \State Store $\mathit{MakeTransitions}(Subnet, S)$ in $l$;
	\EndFor
	\State $\overline{comb} = \mathit{Combining}(l)$;
	\For{ each $sv \in SV$ and each $comb \in \overline{comb}$ }
           \State $ot = \mathit{Matching}(sv, \overline{\mathit{comb}}, \overline{\mathit{hole}})$;
           \IfThen{($ot$ is defined)}{Store $ot$ in $L$; ~~~ $\setminus \setminus$ if $\mathit{Matching}()$ succeeds}
	\EndFor
\State \Return $L$;
\end{algorithmic}  
\end{algorithm}

@@

\def\inactive{\{-\}}
\paragraph{Combining.}
The combining method enumerates all the possible behaviours of the
subnets as all the possible combinations of their open transitions.
Assume that there is a collection of $n$ subnets.
%% \noteSB{These variables are not used and conflict with the returned
%%   set $L$ in Alg.~\ref{alg2}, on one hand, and control states $s_i$ of
%%   the subnets, on the other.}
%% \st{$L = {s_1, s_2, ...,  s_n}$;}
We denote $\overline{ot}_i$ the set of open
transitions of the $i$-th
subnet (obtained in line~3 of the algorithm); %% $s_i$ and 
``$-$'' means that the subnet is not involved.
The combination $\overline{comb}$, a set of n-tuples, is the
cartesian product:  
$\overline{comb} = (\inactive\cup \overline{ot}_1) \times (\inactive\cup \overline{ot}_2)\times \dots \times (\inactive\cup \overline{ot}_n)$.

@@

\paragraph{Matching.}
The \emph{Matching} method builds the OTs of a pNet node from those of
its subnets (see rule Tr2 in App.~\ref{appendix:semRules}).
For each synchronisation vector and each possible
combination of behaviours of the subnets, as generated by the \emph
{Combining} method, it builds the corresponding open transition.
Here, we only detail the construction of the predicate.
From a synchronization vector
$sv = \left((a'_i)^{i\in I} (b'_j)^{j\in J}\rightarrow v'\right) \in SV$
and its guard $G_k$; a tuple of open transitions
$C = ({ot}_i)^{i \in [1,n]} \in \overline{comb}$, such that, for each $i \in [1,n]$, either $ot_i = -$, or the result action of $ot_i$ is $a_i$; the hole behaviors $\mathit{Hole} = (b_j)^{j \in J}$; and a fresh variable $v$, representing the result action of the OT under construction, we build the predicate:
%
\begin{multline*}
  \mathit{MkPred}(sv, C, \overline{\mathit{hole}}) = 
  \\
  (\forall i\in I, a_i=a'_i) \land
  (\forall i \not\in I, ot_i = -) \land
  (\forall j\in J, b_j=b'_j) \land
  (v=v') \land G_k.
\end{multline*}

@@

\begin{figure}[t]
\begin{mathpar}
    ot = \openrule{\{s0 \xrightarrow{fail(true)} s0, t0 \xrightarrow{resume(false)} t0\}, ~~\{\xrightarrow{hb}\},~~fail(true)=fail(b_1)\\
    \OTland resume(false)=start(b_2)\OTland hb=fail(b_0)\OTland v=\underline{fail}\OTland b_1=b_2\\
    \OTland (b_1\lor b_2) \Rightarrow b_0, ~~\{\}}
    	   {\ostate{s0,t0} \xrightarrow{v} \ostate{s0,t0}}
  \end{mathpar}
\caption{One of the unsatisfiable open transitions in the Failure Monitor pNet}  \label{schema:unsat-ot}
\end{figure}

@@

\begin{figure}[t]
  \centerline{\includegraphics[width=12cm]{XFIG/FailureTimerOA}}
  \caption{Open Automaton for the Failure Monitor architecture}
  \label{schema:resultOA}
\end{figure}

@@

In this paper, we presented an algorithm for generating an open
automaton representing the semantics of an open pNet.
We have also presented our implementation, which comprises two main parts. Firstly, we
compute all the possible open transitions.
The algorithm combines the actions of subnets and
hole behaviors, which are then matched with the
synchronization vectors by the generated
predicate.  Some of the open transitions obtained at this
intermediate step, constructed in a structural manner from all
possible combinations of possible logical predicates, may
contain predicates which do not represent any possible concrete
instantiations. 
Secondly, to get rid of these useless transitions, we use the SMT solver Z3 for checking the
satisfiability of the predicate in each open transition. 
To this end, we encode into Z3 the representations of the action algebra and
the predicates before submitting them to the
Z3 solver. We implemented our algorithm in the VerCors platform and 
tested the algorithm on available examples, such as encoding expressions from various process algebras. Our case-studies show that our encoding
successfully identifies all unsatisfiable open transitions and that
the resulting automata correctly reflect the expected 
movements of the encoded process expressions.
